Adversarial Training of Neural Networks\\ against Systematic Uncertainty

### 2016-08-02 paper.tex
method

Let assume a probability space $(\Omega, {\cal F}, P)$, where $\Omega$ is a
sample space, ${\cal F}$ is a set of events and $P$ is a probability measure.
Let consider the multivariate random variables $X_\lambda: \Omega \mapsto
\mathbb{R}^p$ and $Y: \Omega \mapsto {\cal Y}$, where $X_\lambda$ depends on a
nuisance parameter $\lambda$ whose values define the family of its systematic uncertainties.
That is, $X_\lambda$ and $Y$ induce together
a joint probability distribution $p(X,Y|\lambda)$, where the
conditional on $\lambda$ denotes $X_\lambda$. For training, let further assume a
finite set $\{ x_i, y_i, \lambda_i \}_{i=1}^N$ of realizations
$X_{\lambda_i}(\omega_i), Y(\omega_i)$, for $\omega_i \in \Omega$ and known
values $\lambda_i$ of the nuisance parameter. Our goal is to learn a function $f :
\mathbb{R}^p \mapsto {\cal Y}$ (e.g., a classifier if ${\cal Y}$ is a finite set
of classes) minimizing the expected value of a loss $L(Y, f(X_{\lambda}))$, with
the constraint that $f(X_\lambda)$ should be robust to the value of the nuisance
parameter $\lambda$ -- which remains unknown at test time. More specifically, we
aim at building $f$ such that in the ideal case
\begin{equation}\label{eqn:criterion-true}
f(X_{\lambda_i}(\omega)) = f(X_{\lambda_j}(\omega))
\end{equation} for any
sample $\omega \in \Omega$ and any $\lambda_i, \lambda_j$ pair of values of the
nuisance parameter.

Since we do not have training tuples $(X_{\lambda_i}(\omega),
X_{\lambda_j}(\omega))$ (for the same unknown $\omega$), we propose instead to
solve the closely related problem of finding a predictive function $f$ such that
\begin{equation}\label{eqn:criterion}
    P(\{ \omega | f(X_{\lambda_i}(\omega)) = y \} ) = P( \{ \omega' | f(X_{\lambda_j}(\omega')) = y\}) \text{ for all $y \in {\cal Y}$}.
\end{equation}
In words, we are looking for a predictive function $f$ such that  the
distribution of $f(X_\lambda)$ is invariant with respect to the nuisance
parameter $\lambda$. Note that a function $f$ for which Eqn.~\ref{eqn:criterion-true} is
true necessarily satisfies Eqn.~\ref{eqn:criterion}. The converse is however in
general not true, since the sets of samples $\{ \omega | f(X_{\lambda_i}(\omega)) = y \}$
and $\{ \omega' | f(X_{\lambda_j}(\omega')) = y \}$ do not need to be the same
for the equality to hold.

@@

Adversarial training was first proposed by \cite{goodfellow2014generative} as a
way to build a generative model capable of producing samples from random noise
$z \sim p_Z$. More specifically, the authors pit a generative model $G$ against
an adversary classifier $D$ whose repelling objective is to recognize real from
generated data. Both models $G$ and $D$ are trained simultaneously, in such a
way that $G$ learns to produce samples that are difficult to identify by $D$,
while $D$ incrementally adapts to changes in $G$. At the equilibrium, $G$ models
a distribution whose samples can be identified by $D$ only by chance. In other
words, assuming enough capacity in $D$ and  $G$, the distribution $p_{G(Z)}$
eventually converges towards the real distribution $p_X$.

In this work, we repurpose adversarial training for regularizing the
construction of the predictive model $f$ so as to satisfy
Eqn.~\ref{eqn:criterion}.

\glnote{describe baseline}
\glnote{describe adversarial approach}
\glnote{proof that it solves Eqn. 2}

>>>>

Let assume a probability space $(\Omega, {\cal F}, P)$, where $\Omega$ is a
sample space, ${\cal F}$ is a set of events and $P$ is a probability measure.
Let consider the multivariate random variables $X_\lambda: \Omega \mapsto
\mathbb{R}^p$ and $Y: \Omega \mapsto {\cal Y}$, where $X_\lambda$ depends on a
nuisance parameter $\lambda \in \mathbb{R}$ whose values define a continuously
parameterized family of its systematic uncertainties. That is, $X_\lambda$ and
$Y$ induce together a joint probability distribution $p(X,Y|\lambda)$, where the
conditional on $\lambda$ denotes $X_\lambda$. For training, let further assume a
finite set $\{ x_i, y_i, \lambda_i \}_{i=1}^N$ of realizations
$X_{\lambda_i}(\omega_i), Y(\omega_i)$, for $\omega_i \in \Omega$ and known
values $\lambda_i$ of the nuisance parameter. Our goal is to learn a function
$f(\cdot;\theta_f) : \mathbb{R}^p \mapsto {\cal Y}$ of parameters $\theta_f$
(e.g., a neural network-based classifier if ${\cal Y}$ is a finite set of
classes) minimizing  a loss ${\cal L}_f(\theta_f)$, with the constraint that $f(X_\lambda ;
\theta_f)$ should be robust to the value of the nuisance parameter $\lambda$ --
which remains unknown at test time. More specifically, we aim at building $f$
such that in the ideal case
\begin{equation}\label{eqn:criterion-true}
f(X_{\lambda_i}(\omega) ; \theta_f) = f(X_{\lambda_j}(\omega) ; \theta_f)
\end{equation} for any
sample $\omega \in \Omega$ and any $\lambda_i, \lambda_j$ pair of values of the
nuisance parameter.

Since we do not have training tuples $(X_{\lambda_i}(\omega),
X_{\lambda_j}(\omega))$ (for the same unknown $\omega$), we propose instead to
solve the closely related problem of finding a predictive function $f$ such that
\begin{equation}\label{eqn:criterion}
    P(\{ \omega | f(X_{\lambda_i}(\omega) ; \theta_f) = y \} ) = P( \{ \omega' | f(X_{\lambda_j}(\omega') ; \theta_f) = y\}) \text{ for all $y \in {\cal Y}$}.
\end{equation}
In words, we are looking for a predictive function $f$ such that  the
distribution of $f(X_\lambda; \theta_f)$ is invariant with respect to the nuisance
parameter $\lambda$. Note that a function $f$ for which Eqn.~\ref{eqn:criterion-true} is
true necessarily satisfies Eqn.~\ref{eqn:criterion}. The converse is however in
general not true, since the sets of samples $\{ \omega | f(X_{\lambda_i}(\omega); \theta_f) = y \}$
and $\{ \omega' | f(X_{\lambda_j}(\omega'); \theta_f) = y \}$ do not need to be the same
for the equality to hold.

@@

Adversarial training was first proposed by \cite{goodfellow2014generative} as a
way to build a generative model capable of producing samples from random noise
$z \sim p_Z$. More specifically, the authors pit a generative model $g: {\cal Z}
\mapsto \mathbb{R}^p$ against an adversary classifier $d : \mathbb{R}^p \mapsto \{
0, 1\}$ whose antagonistic objective is to recognize real data $X$ from generated data $g(Z)$. Both
models $g$ and $d$ are trained simultaneously, in such a way that $g$ learns to
produce samples that are difficult to identify by $d$, while $d$ incrementally
adapts to changes in $g$. At the equilibrium, $g$ models a distribution whose
samples can be identified by $d$ only by chance. In other words, assuming enough
capacity in $d$ and  $g$, the distribution $p_{g(Z)}$ eventually converges
towards the real distribution $p_X$.

In this work, we repurpose adversarial training as a means to regularize the
predictive model $f$ in order to satisfy Eqn.~\ref{eqn:criterion}. In
particular, we pit $f$ against an adversary regression model $r(\cdot ;
\theta_r) : \mathbb{R} \mapsto \mathbb{R}$ of parameters $\theta_r$ and
associated loss ${\cal L}_r(\theta_r)$, taking as input realizations of $f(X_\lambda; \theta_f)$
and producing as output a prediction $\smash{\hat\lambda} = r(f(X_\lambda); \theta_f)$
of the nuisance parameter. If $f(X_\lambda; \theta_f)$ varies with $\lambda$,
then the corresponding correlation can be captured by $r$. By contrast, if
$f(X_\lambda; \theta_f)$ is invariant with $\lambda$ as we require, then $r$
should perform poorly. Training $f$ such that it minimizes the performance of $r$
should therefore acts as a regularization towards Eqn.~\ref{eqn:criterion}.

As for generative adversarial networks, we propose to
train $f$ and $r$ simultaneously, which can be carried out by considering
the value function
\begin{equation}
    E(\theta_f, \theta_r) = {\cal L}_f(\theta_f) - {\cal L}_r(\theta_r)
\end{equation}
and finding the saddle point $\smash{\hat\theta_f}, \smash{\hat\theta_r}$ such that
\begin{align}
    \smash{\hat\theta_f} &= \arg \min_{\theta_f} E(\theta_f, \smash{\hat\theta_r}), \\
    \smash{\hat\theta_r} &= \arg \max_{\theta_r} E(\smash{\hat\theta_f}, \theta_r).
\end{align}

\glnote{What are the necessary conditions on ${\cal L}_r$ to imply
Eqn.~\ref{eqn:criterion} at the saddle point? We should clarify and prove that
formally! There may also be assumptions required on the learning problem itself.}
